---
pdf_document:
  number_sections: yes
  toc: yes
  tocdepth: 2
author: "Daria Svidzinska"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
title: "Description of Surface Metrics"
fontsize: 12pt
bibliography: surface_metrics_references.bib
link-citations: yes
linkcolor: blue
csl: apa.csl
# it's imporatnt when you knit to make sure that you knit to project directory
# Knit -> Knit Directory -> Project Directory
# https://stirlingcodingclub.github.io/Manuscripts_in_Rmarkdown/Rmarkdown_notes.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Landscape metrics implemented in FragStats have been a rule of thumb in the analysis of landscape pattern for several reasons. First of all, it was a useful tool to quantify landscape patch and pattern structure as it was presented in all landscape ecological studies. Secondly, there is a huge set a well-defined quantitative metrics available for any task whether it is landscape diversity, complexity or simply fragmentation. And last but not least there is a specially designed and well-documented software package available for free.

But @McGarigal2009 suggested that a classical view of landscape patch-and-matrix matrix is too simplified comparing to real-world continuity. They suggested that discrete models should be replaced (or at least try to be) by a gradient model. And while at first it was pretty difficult to find and represent landscape gradients, with the invention of advanced geoinformational data representation and storage capabilities, as well as billions of terabytes of remote sensing data continuously becoming available for free it became more clear that it is much easier and appropriate to use gradient model to represent and analyze landscape and its properties.

# Training Data

For illustration I am using two datasets from the ./data folder. The first one is ALOS DEM with the spatial resolution of 30 m. Both datasets were standardized by substracting mean values and dividing to standard deviation.

```{r, echo = FALSE, eval = TRUE, message = FALSE, fig.width = 4, fig.height = 2.5, fig.fullwidth = TRUE, fig.cap = "Digital elevation model"}
par(mar = c(2, 2, 0.2, 0.2)) # bottom, left, top, right
library(raster)
dem <- raster("./data/dem.tif")
dem_min <- cellStats(dem, stat = "min")
dem_max <- cellStats(dem, stat = "max")
dem_norm <- (dem - dem_min)/(dem_max - dem_min)
plot(dem_norm, col = terrain.colors(255))

```

The second one is Sentinel-derived NDVI as of 22-07-2017 with the spatial resolution 10 m.

```{r, echo = FALSE, eval = TRUE, message = FALSE, fig.width = 4, fig.height = 2.5, fig.fullwidth = TRUE, fig.cap = "NDVI 22-07-2017"}
par(mar = c(2, 2, 0.2, 0.2)) # bottom, left, top, right
library(raster)
ndvi <- raster("./data/ndvi.tif")
ndvi_min <- cellStats(ndvi, stat = "min")
ndvi_max <- cellStats(ndvi, stat = "max")
ndvi_norm <- (ndvi - ndvi_min)/(ndvi_max - ndvi_min)
plot(ndvi_norm)

```


# Surface Metrics

## Amplitude Metrics

Amplitude Metrics measure vertical characteristics of surface deviations. These metrics are sensitive to variability in the overall height distribution, but not the spatial arrangement, location or distribution of surface peaks and valleys. As such, these parameters measure aspects of landscape composition, not configuration.

### Average Rougness (Arithmetic Average Height aka Centre Line Average)

The average absolute deviation of the surface heights from the mean. This is a general measure of overall surface variability and can be interpreted as a nonspatial measure of landscape diversity, analogous to the patch-based diversity metrics. Larger values represent an increasing range of values in the surface attribute (akin to increasing patch richness) and/or an increasing spread in the distribution of area among levels (heights) of the surface attribute (akin to increasing patch evenness). Importantly, this metric does not differentiate among different shapes of the surface height profile. 

$$R_a = \frac{1}{n} \sum_{i=1}^{n} \left| \bar{z} - z_i \right|$$

where $n$ - number of pixels in a given patch; $\bar{z}$ - mean value; $z_i$ - the value of the $i$^th^ pixel in a given patch. It approximates surface roughness by calculating the mean absolute departure of the surface values from the mean plane.


```{r, echo = FALSE, eval = TRUE, message = FALSE, fig.width = 4, fig.height = 2.5, fig.fullwidth = TRUE, fig.cap = "DEM surface average roughness"}
par(mar = c(2, 2, 0.2, 0.2)) # bottom, left, top, right
library(raster)
dem <- raster("./data/dem.tif")
dem_min <- cellStats(dem, stat = "min")
dem_max <- cellStats(dem, stat = "max")
dem_norm <- (dem - dem_min)/(dem_max - dem_min)
mean <- cellStats(dem_norm, stat = "mean")
surf_rough <- abs(mean - dem_norm)
plot(surf_rough, col = gray.colors(255))

```

```{r, echo = FALSE, eval = TRUE, message = FALSE, fig.width = 4, fig.height = 2.5, fig.fullwidth = TRUE, fig.cap = "NDVI surface average roughness"}
par(mar = c(2, 2, 0.2, 0.2)) # bottom, left, top, right
library(raster)
ndvi <- raster("./data/ndvi.tif")
ndvi_min <- cellStats(ndvi, stat = "min")
ndvi_max <- cellStats(ndvi, stat = "max")
ndvi_norm <- (ndvi - ndvi_min)/(ndvi_max - ndvi_min)
mean <- cellStats(ndvi_norm, stat = "mean")
surf_rough <- abs(mean - ndvi_norm)
plot(surf_rough, col = gray.colors(255))

```

### Root mean square roughness 

The standard deviation of the distribution of surface heights. This is a general measure of overall surface variability like $R_a$, but it is more sensitive than $R_a$ to large deviations from the surface mean. Otherwise, this metric has the same general interpretation as $R_a$ and is likely to be highly correlated with $R_a$ in real-world applications.

$$R_q = \sqrt{\frac{1}{n} \sum_{i=1}^{n} \left| \bar{z} - z_i \right|^2}$$

This parameter is also known as RMS. The RMS mean line is the line that divides the profile so that the sum of the squares of the deviations of the profile height from it is equal to zero.

```{r, echo = FALSE, eval = TRUE, message = FALSE, fig.width = 4, fig.height = 2.5, fig.fullwidth = TRUE, fig.cap = "DEM surface root mean square roughness"}
par(mar = c(2, 2, 0.2, 0.2)) # bottom, left, top, right
library(raster)
dem <- raster("./data/dem.tif")
dem_min <- cellStats(dem, stat = "min")
dem_max <- cellStats(dem, stat = "max")
dem_norm <- (dem - dem_min)/(dem_max - dem_min)
mean <- cellStats(dem_norm, stat = "mean")
rms_rough <- sqrt(abs(mean - dem_norm)^2)
plot(rms_rough, col = gray.colors(255))

```

```{r, echo = FALSE, eval = TRUE, message = FALSE, fig.width = 4, fig.height = 2.5, fig.fullwidth = TRUE, fig.cap = "NDVI surface root mean square roughness"}
par(mar = c(2, 2, 0.2, 0.2)) # bottom, left, top, right
library(raster)
ndvi <- raster("./data/ndvi.tif")
ndvi_min <- cellStats(ndvi, stat = "min")
ndvi_max <- cellStats(ndvi, stat = "max")
ndvi_norm <- (ndvi - ndvi_min)/(ndvi_max - ndvi_min)
mean <- cellStats(ndvi_norm, stat = "mean")
rms_rough <- sqrt(abs(mean - ndvi_norm)^2)
plot(rms_rough, col = gray.colors(255))

```

# References

